---
title: "Solar PV Suitability Analysis: User Guide"
date: 2026-07-15
author: Kenzie Thomson
format: html
toc: true
toc-location: left
toc-depth: 4
theme: minty
title-block-banner: images/solarfarm.jpg
title-block-banner-color: white
---

# About This Project

As electricity demand grows across British Columbia, more renewable energy installations need to be developed to support a resilient, diversified, low-carbon future. In light of this challenge, a site suitability analysis was undergone to create a province-wide screening tool that determines the most optimal locations for solar PV development.

This analysis is broken down into 2 main sections:

**Section 1** determines the geographical potential of BC - the total land area where solar farm development is possible.

**Section 2** determines the technical potential of BC - the total energy these sites could produce on a yearly basis.

By providing clear and accessible information on renewable energy potential, this project helps accelerate clean energy development, informs decision-making and supports BC's transformation to a more resilient and sustainable electricity system.

This document provides a detailed guide of the analysis, including all code used to derive results.

[Click for more info](https://www.uvic.ca/acet/projects/bc-renewable-energy-mapping/index.php)

[Click for first-draft publication](https://borealisdata.ca/dataset.xhtml?persistentId=doi:10.5683/SP3/MZYJFF)

[Click for raw data download](tbd)

**NOTE:** You will need to change each file path to a folder on your computer!

# Set-Up

First, each package must be installed and loaded to complete this analysis

```{r}
#| output: false
#| message: false

# Load packages
library(terra)
library(sf)
library(tidyverse)
library(exactextractr)
library(AHPtools) 
library(ggplot2)
library(scales) 
library(dplyr)
library(classInt)
library(bcmaps)
library(stringr)
```

I recommend setting up an organized file system, similar to the one below, to save your intermediate outputs and final results.

```text
Solar_Suitability/
├── raw_data/
├── criteria/
│   ├── cleaned/         # Data layers reprojected, clipped, masked, resampled, and rasterized
│   ├── reclassified/    # Data layers reclassified to a 1-5 suitability scale
│   ├── slope/           # DEM-derived slope pieces
│   └── aspect/          # DEM-derived aspect pieces
├── constraints/
│   ├── binary/          # Data layers brought to a binary classification
│   └── mask/            # Final constraints mask
└── suitability_surfaces/
    ├── unmasked/        # Suitability results without constraints mask
    └── masked/          # Suitability results with constraints mask
```‎

# Stage 1: Geographical Potential

## Step 1: Data Pre-Processing

### 1.1 Create Template

To first step of the analysis is the creation of a template raster. This layer will ensure that every input layer is the exact same extent and resolution. We will do create this by rasterizing a polygon representing BCs administrative boundary.

```{r}
#| message: false
#| eval: false

# Create template  

# Read in layer
bc <- sf::st_read("~/Library/CloudStorage/OneDrive-UBC/RA/Data/Admin Boundaries/BC_RegionalDistricts.shp") 

# Project to BC Enviro Albers (EPSG:3005)
bc <- sf::st_transform(bc, crs = 3005)

# Create template spanning vector extent
# Set 30 m spatial resolution (each cell = 30m x 30m) 
template <- rast(bc, res = 30)

# Convert to raster and burning in the Regional District Name attribute column 
template <- rasterize(bc, template, field = "NAME_2")

# Save the output
writeRaster(template, "~/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/template/template.tif", overwrite = TRUE)

# Quick read
template <- rast("~/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/template/template.tif")
```

The next step of the analysis is data pre-processing. This section is divided in 2 parts - suitability criteria and land constraints.

### 1.2 Suitability Criteria Processing

There are 6 suitability criteria layers used in this analysis:

1.  Global Tilted Irradiance (GTI)

2.  Ground Slope

3.  Ground Aspect

4.  Proximity to Substations

5.  Proximity to Demand Centers

6.  Proximity to Major Roads

Each layer needs to be clipped to BC's bounding box, masked to a template raster, projected to a common CRS (EPSG:3005 - BC Environmental Albers), sampled to a consistent spatial resolution, and reclassified to a 1-5 suitability scale. We will process each layer individually to ease computational power.

#### 1.2.1 Global Tilted Irradiance

Data source: [Global Solar Atlas GIS Data](https://globalsolaratlas.info/download/canada)

Clean data:

```{r}
#| message: false 
#| eval: false

# Read in GTI raster layer 
GTI <- rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Existing Renewables Mapping Project/Raw solar and wind/Canada_GISdata_LTAy_YearlyMonthlyTotals_GlobalSolarAtlas-v2_GEOTIFF/GTI.tif")

# Project to template raster projection (EPSG:3005)
# Resampling method = bilinear interpolation (used for continuous data)
gti_proj <- project(GTI, template, method = "bilinear")

# Clip to BC bounding box
gti_crop <- crop(gti_proj, template)

# Mask to BC administrative boundary
gti_mask <- mask(gti_crop, template) 

# Save in cleaned data folder
writeRaster(gti_mask, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/cleaned/gti_cleaned.tif", overwrite = TRUE)
```

Reclassify to 1-5 suitability scale:

```{r}
#| message: false
#| eval: false

# Reclassify using natural breaks (jenks) 

# Get data bounds
true_min <- global(gti_mask, "min", na.rm = TRUE)[1, 1]
true_max <- global(gti_mask, "max", na.rm = TRUE)[1, 1]

set.seed(123)
vals <- terra::spatSample(gti_mask, 500000, method = "regular", na.rm = TRUE)
jenks_obj <- classIntervals(as.numeric(vals[[1]]), n = 5, style = "fisher") 
breaks <- jenks_obj$brks

# Set data bounds 
breaks[1] <- true_min
breaks[length(breaks)] <- true_max 

# Create classification matrix
m_jenks <- matrix(c(
  head(breaks, -1),
  tail(breaks, -1),
  1:5
), ncol = 3)

# Execute classification 
gti_reclass <- terra::classify(gti_mask, m_jenks, include.lowest = TRUE)

# Save in reclassified data folder
writeRaster(gti_reclass, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/gti_reclassified.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Visualize Criteria 1 - GTI"
suitability_cols <- c("#FFFFCC", "#FECC5C","#FD8D3C","#F03B20","#BD0026")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/gti_reclassified.tif"), col=suitability_cols)
```

#### 1.2.2 Ground Slope

Data source: [bcmaps R Package](https://cran.r-project.org/web/packages/bcmaps/index.html)

The ground slope raster is created by piecing together a DEM from the bcmaps package and then utilizing terrain functions to calculate slope. Initiate this for-loop to download the DEM, create the slope raster, and mask to our template raster. Expect this code chunk to take a few hours to finish.

```{r}
#| message: false
#| eval: false

# Check available layers in bcmaps package
bc_maps_avail <- available_layers()

# Download DEM by ecoregion, check layer names
ecoprov <- ecoprovinces() 

# Set output directory
out_dir <- "Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/slope"

# Define ecoprovinces to process in a list
ecoprovs <- c(
  "SAL", "NBM", "TAP", "BOP", "SBI",
  "SIM", "SOI", "COM", "GED", "NEP", "CEI"
)

# Call template raster mask template
template <- rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/template/template.tif") 

# Initiate for-loop to create slope raster from DEM chunks 
for (prov in ecoprovs) {
  
  message("Processing ", prov)
  
  # Get ecoprov boundary
  bound_sf <- ecoprovinces() %>%
    filter(ECOPROVINCE_CODE == prov)
  
  # Convert to SpatVector and project to BC Albers (EPSG:3005)
  bound_vect <- vect(bound_sf)
  bound_vect <- project(bound_vect, "EPSG:3005")
  
  # BUFFER: Create a 500m buffer around the boundary to pull neighboring DEM data
  bound_buffered <- buffer(bound_vect, width = 500) 
  
  # Get DEM using the buffered boundary 
  dem <- cded_raster(sf::st_as_sf(bound_buffered))
  
  # Project DEM to template grid
  dem_3005 <- project(
    rast(dem),
    template,
    method = "bilinear")
  
  # Calculate slope on the buffered DEM first to ensure seamless edges
  slope <- terrain(
    dem_3005,
    v = "slope",
    neighbors = 8,
    unit = "degrees")
  
  # Crop and mask to the strict, original ecoprovince boundary
  slope_clipped <- crop(slope, bound_vect)
  slope_clipped <- mask(slope_clipped, bound_vect)
  
  # Write slope raster
  writeRaster(slope_clipped, file.path(out_dir, paste0(prov, "_slope.tif")), overwrite = TRUE)
  
  # Clean memory
  rm(dem, dem_3005, slope, slope_clipped, bound_vect, bound_buffered)
  gc()
}

# Now join all individual ecoprovince slope rasters together 

# List and load slope rasters
slope_files <- list.files(path = out_dir, pattern = "_slope\\.tif$", full.names = TRUE)
slope_list <- lapply(slope_files, rast)

# Merge all slopes into BC-wide raster
bc_slope <- do.call(merge, slope_list)

# Force exact match to template extent, then mask
bc_slope_aligned <- crop(bc_slope, template)
bc_slope_final <- mask(bc_slope_aligned, template)

# Save in cleaned data folder
writeRaster(bc_slope_final, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/cleaned/bc_slope_cleaned.tif")
```

Reclassify to a 1-5 suitability scale:

```{r}
#| message: false
#| eval: false

# Reclassify using a defined matrix

# Create classes using format (from, to, new)
m <- c(0, 5, 5, #best
       5, 10, 4, 
       10, 15, 3, 
       15, 20, 2, 
       20, 100, 1) #worst 

# Create matrix
slope_matrix <- matrix(m, ncol = 3, byrow = TRUE)

# Apply reclassification 
slope_reclass <- terra::classify(bc_slope_final, slope_matrix, include.lowest = TRUE, right = FALSE)

# Save in reclassified data folder
writeRaster(slope_reclass, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/slope_reclassified.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Visualize Criteria 2 - Ground Slope"
suitability_cols <- c("#FFFFCC", "#FECC5C","#FD8D3C","#F03B20","#BD0026")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/slope_reclassified.tif"), col=suitability_cols)
```

#### 1.2.3 Ground Aspect

Data source: [bcmaps R Package](https://cran.r-project.org/web/packages/bcmaps/index.html)

We will apply the same for-loop methodology to create a BC wide aspect layer from a DEM. Again, expect this code chunk to take a few hours to finish.

```{r}
#| message: false
#| eval: false

# Check available layers in bcmaps package
bc_maps_avail <- available_layers()

# Download DEM by ecoregion, check layer names
ecoprov <- ecoprovinces() 

# Set output directory
out_dir <- "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/aspect"

# Define ecoprovinces to process in a list
ecoprovs <- c(
  "SAL", "NBM", "TAP", "BOP", "SBI",
  "SIM", "SOI", "COM", "GED", "NEP", "CEI"
)

# Call template raster mask template
template <- rast("~/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/template/template.tif") 

# Initiate loop
for (prov in ecoprovs) {
  
  message("Processing ", prov)
  
  # Get ecoprov boundary
  bound_sf <- ecoprovinces() %>%
    filter(ECOPROVINCE_CODE == prov)
  
  bound_vect <- vect(bound_sf)
  bound_vect <- project(bound_vect, "EPSG:3005")
  
  # BUFFER: Create a 500m buffer to pull overlapping DEM data
  bound_buffered <- buffer(bound_vect, width = 500)
  
  # Get DEM using the buffered boundary
  dem <- cded_raster(sf::st_as_sf(bound_buffered))
  
  # Project DEM to template
  dem_3005 <- project(
    rast(dem),
    template,
    method = "bilinear")
  
  # Calculate aspect on the buffered DEM
  aspect <- terrain(
    dem_3005,
    v = "aspect",
    neighbors = 8,
    unit = "degrees")
  
  # Crop and mask strictly to the original ecoprovince boundary
  aspect_clipped <- crop(aspect, bound_vect)
  aspect_clipped <- mask(aspect_clipped, bound_vect)
  
  # Write aspect raster
  writeRaster(
    aspect_clipped,
    file.path(out_dir, paste0(prov, "_aspect.tif")),
    overwrite = TRUE)
  
  # Clean memory
  rm(dem, dem_3005, aspect, aspect_clipped, bound_vect, bound_buffered)
  gc()
}

# Now join all individual ecoprovince aspect rasters together 

# List all aspect rasters
aspect_files <- list.files(
  path = out_dir,
  pattern = "_aspect\\.tif$",
  full.names = TRUE)

# Load rasters into a list
aspect_list <- lapply(aspect_files, rast)

# Merge into one surface
bc_aspect <- do.call(merge, aspect_list)

# Crop and mask to template 
bc_aspect_aligned <- crop(bc_aspect, template)
bc_aspect_final <- mask(bc_aspect_aligned, template)

# Save in cleaned data folder
writeRaster(bc_aspect_final, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/cleaned/bc_aspect_cleaned.tif", overwrite = TRUE)
```

Reclassify to a 1-5 suitability scale:

```{r}
#| message: false
#| eval: false

# Reclassify using a defined matrix

# Create classes using format (from, to, new)
m <- c(315, 360, 1, #N
       0, 45, 1, #N (wrap-around)
       45, 135, 3, #E
       135, 225, 5, #S
       225, 315, 3) #W

# Create matrix
aspect_matrix <- matrix(m, ncol = 3, byrow = TRUE)

# Apply classification
aspect_reclass <- terra::classify(bc_aspect_final, aspect_matrix, include.lowest = TRUE, right = TRUE)

# Save in reclassified data folder
writeRaster(aspect_reclass, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/aspect_reclassified.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Visualize Criteria 3 - Ground Aspect"
suitability_cols <- c("#FFFFCC", "#FECC5C","#FD8D3C","#F03B20","#BD0026")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/aspect_reclassified.tif"), col=suitability_cols)
```

#### 1.2.4 Proximity to Substations

Data source: Private

Substation locations are point data, therefor, we must first create a euclidean distance raster where each cell represents a distance in meters away from the closest substation.

```{r}
#| message: false
#| eval: false

# Load in vector layer
subs <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/Substations/substations.shp")

# Transform to spatvect for processing
subs_v <- vect(subs)

# Create Euclidean distance raster showing proximity to substations that spatially conforms to our template
subs_dist <- terra::distance(template, subs_v, rasterize = TRUE)

# Mask to template boundary
subs_mask <- mask(subs_dist, template)

# Save in cleaned data folder
writeRaster(subs_mask, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/cleaned/subs_cleaned.tif", overwrite = TRUE)
```

Reclassify to a 1-5 suitability scale:

```{r}
#| message: false
#| eval: false

# Reclassify using a defined matrix

# Units = meters because our crs = projected to BC Albers
m <- c(0, 1000, 5, # 1km surrounding substations is best
       1000, 2000, 4,
       2000, 5000, 3,
       5000, 7500, 2,
       7500, Inf, 1) # Anything beyond 7.5km is worst 

# Create matrix
subs_m <- matrix(m, ncol = 3, byrow = TRUE)

# Apply classification
subs_reclass <- terra::classify(subs_mask, subs_m, include.lowest = TRUE, right = FALSE)

# Save in reclassified data folder
writeRaster(subs_reclass, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/subs_reclassified.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Visualize Criteria 4 - Proximity to Substations"
suitability_cols <- c("#FFFFCC", "#FECC5C","#FD8D3C","#F03B20","#BD0026")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/subs_reclassified.tif"), col=suitability_cols)
```

#### 1.2.5 Proximity to Demand Centers

Data source: [BC Data Catalogue](https://catalogue.data.gov.bc.ca/dataset/bc-major-cities-points-1-2-000-000-digital-baseline-mapping)

Repeat process using point locations of major cities.

```{r}
#| message: false
#| eval: false

# Load in vector layer
cities <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/DBM_BC_7H_MIL_POPULATION_POINT/DBMBC7HML4_point.shp")

# Transform to spatvect for processing
cities_v <- vect(cities)

# Create Euclidean distance raster showing proximity to substations that spatially conforms to our template
cities_dist <- terra::distance(template, cities_v, rasterize = TRUE)

# Mask to template boundary
cities_mask <- mask(cities_dist, template)

# Save in cleaned data folder
writeRaster(cities_mask, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/cleaned/cities_cleaned.tif", overwrite = TRUE)
```

Reclassify to a 1-5 suitability scale:

```{r}
#| message: false
#| eval: false

# Reclassify using a defined matrix

# Units = meters because our crs = projected to BC Albers
m <- c(0, 5000, 5, 
       5000, 15000, 4,
       15000, 30000, 3,
       30000, 45000, 2,
       45000, Inf, 1)  

# Create matrix
cities_m <- matrix(m, ncol = 3, byrow = TRUE)

# Apply classification
cities_reclass <- terra::classify(cities_mask, cities_m, include.lowest = TRUE, right = FALSE)

# Save in reclassified data folder
writeRaster(cities_reclass, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/cities_reclassified.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Visualize Criteria 5 - Proximity to Demand Centers"
suitability_cols <- c("#FFFFCC", "#FECC5C","#FD8D3C","#F03B20","#BD0026")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/cities_reclassified.tif"), col=suitability_cols)
```

#### 1.2.6 Proximity to Major Roads

Data source: [BC Data Catalogue](https://catalogue.data.gov.bc.ca/dataset/7-5m-major-roads-the-atlas-of-canada-base-maps)

Repeat process using locations of major roads.

```{r}
#| message: false
#| eval: false

# Load in vector layer 
roads <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/BCGW_02001F02_1784319511692_11424/DBM_BC_7H_MIL_ROADS_LINE.gdb")

# Transform to spatvect for processing
roads_v <- vect(roads)

# Create Euclidean distance raster showing proximity to substations that spatially conforms to our template
roads_dist <- terra::distance(template, roads_v, rasterize = TRUE)

# Mask to template boundary
roads_mask <- mask(roads_dist, template)

# Save in cleaned data folder
writeRaster(roads_mask, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/cleaned/roads_cleaned.tif", overwrite = TRUE)
```

Reclassify to a 1-5 suitability scale:

```{r}
#| message: false
#| eval: false

# Reclassify using a defined matrix

# Units = meters because our crs = projected to BC Albers
m <- c(0, 1000, 5, 
       1000, 5000, 4,
       5000, 15000, 3,
       15000, 30000, 2,
       30000, Inf, 1)  

# Create matrix
roads_m <- matrix(m, ncol = 3, byrow = TRUE)

# Apply classification
roads_reclass <- terra::classify(roads_mask, roads_m, include.lowest = TRUE, right = FALSE)

# Save in reclassified data folder
writeRaster(roads_reclass, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/roads_reclassified.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Visualize Criteria 6 - Proximity to Major Roads"
suitability_cols <- c("#FFFFCC", "#FECC5C","#FD8D3C","#F03B20","#BD0026")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/criteria/reclassified/roads_reclassified.tif"), col=suitability_cols)
```

### 1.3 Land Constraints Processing

This portion of the analysis creates the land constraints spatial mask used to eliminate land from the suitability analysis that is restricted from solar farm development due to technical, environmental, or regulatory factors.

Land constraints include:

1.  Land covers - water, permanent snow and ice, wetlands, artificial and built environments

2.  Linear Features - All roads (including FSRs) and railways

3.  Densely forested areas (canopy density \> 40%)

4.  Agricultural Land Reserve with a Soil Capability Class 1-3 

5.  Protected and conserved areas

6.  Archaeologically and culturally sensitive areas

Each layer will be individually transformed to a binary where:

- 0 = land included in the constraints mask (not suitable)
- 1 = land that is not included in the constraints mask (suitable) 

Each layers cells classified as '0' will then be merged to create a master restriction zone layer.

#### 1.3.1 Land Cover

Data source: [Government Canada Census of Environment Land Cover Register](https://www150.statcan.gc.ca/n1/pub/16-510-x/16-510-x2025002-eng.htm)

Clean data:

```{r}
#| message: false
#| eval: false

# Load in land cover register layer
lcr = rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/LCR_RCT_2020.tif")

# Create a BC boundary layer in the projection of the lcr raster so that we can crop it. We cannot crop it using the template raster yet because the crs' are different and cannot alter the template without affecting the grid. 
target_crs <- crs(lcr)
bc_lcr_boundary <- sf::st_transform(bc, target_crs)

# Crop lcr to bc bounding box
lcr_crop <- crop(lcr, bc_lcr_boundary) 

# Now that the layer is cropped to BC (rather than all of Canada), we can reproject it and mask it to our template
# Method = near - use nearest neighbor as method to preserve integer classification of original layer 
lcr_proj <- project(lcr_crop, template, method = "near")

# Mask to BC administrative boundary
lcr_mask <- mask(lcr_proj, template) 

# Save to cleaned constraints folder
writeRaster(lcr_mask, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/cleaned/lcr_cleaned.tif")
```

Reclassify to 1/0 binary using classification codes listed below:

| Class Name                      | Class Code | Suitable? |
|---------------------------------|------------|-----------|
| Built up and artificial surface | 1          | No        |
| Cropland                        | 2          | Yes       |
| Inland water body               | 3          | No        |
| Treed (non-wetland)             | 4          | Yes       |
| Treed wetland                   | 5          | No        |
| Treed area disturbance          | 6          | Yes       |
| Grassland and shrubland         | 7          | Yes       |
| Wetland (non-treed)             | 8          | No        |
| Sparsely vegetated land         | 9          | Yes       |
| Barren land                     | 10         | Yes       |
| Permanent snow and ice          | 11         | No        |

```{r}
#| message: false
#| eval: false

# Create reclassification scheme 
# 0 = not suitable - within constraints
# 1 = suitable - not within constraints
m <- c(1, 0, 
       2, 1, 
       3, 0, 
       4, 1, 
       5, 0, 
       6, 1, 
       7, 1,
       8, 0, 
       9, 1, 
       10, 1, 
       11, 0) 

# Create matrix
lcr_matrix <- matrix(m, ncol = 2, byrow = TRUE)

# Apply classification
lcr_binary <- terra::classify(lcr_mask, lcr_matrix)

# Save to reclassified folder
writeRaster(lcr_binary, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/lcr_binary.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Constraint 1: Unsuitable Land Covers"
cons_cols <- c("#696969", "#228B22")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/lcr_binary.tif"), col=cons_cols)
```

#### 1.3.2 Linear Features - Roads and Railways

Data source: [BC Data Catalogue](https://catalogue.data.gov.bc.ca/dataset/digital-road-atlas-dra-master-partially-attributed-roads)

Clean the data by selecting the correct layer from the geodatabase, rasterizing it into a binary structure, and masking it to our template.

```{r}
#| message: false
#| eval: false

# Roads

# View available layers in BC Digital Road Atlas geodatabase
st_layers("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/dgtl_road_atlas.gdb")

# DGTL_ROAD_ATLAS_MPAR_SP = includes all the demographic roads in addition to the full dataset of non-demographic roads - roads without names.
# DGTL_ROAD_ATLAS_DPAR_SP = includes only roads that have names. These are primarily highways and local roads, along with named resource roads.

# Select road layer that contains ALL roads
roads <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/dgtl_road_atlas.gdb", layer = "DGTL_ROAD_ATLAS_MPAR_SP") 

# Convert sf object to spatvect
roads_vect <- vect(roads)

# Rasterize layer and spatially conform to our template
# Set field to 0 (unsuitable) and background to 1 (suitable) so all non-road cells are assigned a value of 1
roads_rast <- rasterize(roads_vect, template, field = 0, background = 1)

# Mask to the template 
roads_binary <- mask(roads_rast, template)

# Save into the cleaned constraints folder
writeRaster(roads_binary, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/roads_binary.tif") 
```

```{r}
#| message: false
#| eval: false

# Railways

# Read in railway sf object 
rails <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Mask/Raw Data/railways.gpkg")

# Convert sf object to spatvect
rails_vect <- vect(rails)

# Rasterize layer and spatially conform to our template
# Set field to 0 (unsuitable) and background to 1 (suitable) so all non-road cells are assigned a value of 1
rails_rast <- rasterize(rails_vect, template, field = 0, background = 1)

# Mask to the template 
rails_binary <- mask(rails_rast, template)

# Save into the cleaned constraints folder
writeRaster(rails_binary, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/rails_binary.tif") 
```

```{r}
#| echo: false
#| fig-cap: "Constraint 2: Linear Features - Roads and Railways "

cons_cols <- c("#696969", "#228B22")

r_roads <- rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/roads_binary.tif")
r_rails <- rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/rails_binary.tif")

# Set base layout and tell terra plot not to reset par
par(mfrow = c(1, 2))
plot(r_roads, col = cons_cols, main = "Roads")
plot(r_rails, col = cons_cols, main = "Railways")
```

#### 1.3.3 Dense Forests

Because the land cover register defines treed areas as "land covered with trees with a canopy cover over 10%",  we want to further isolate and protect just those moderate-to-dense forests, which has a canopy cover of 40% (BC Ministry of Forests). 

Data source: [National Terrestrial Ecosystem Monitoring System (NTEMS)](https://opendata.nfis.org/mapserver/nfis-change_eng.html) 

Clean the data by isolating land with a canopy cover greater than 40% by creating a binary and then mask to our template.

```{r}
#| message: false
#| eval: false

# Read in NTEMS canopy cover raster 
canopy <- rast("/Users/kenziethomson/Downloads/CA_canopy_cover_2022/CA_canopy_cover_2022.tif")

#rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Criteria/Raw/CA_canopy_cover_2022.tif")

# Create a BC boundary layer in the projection of the canopy raster so that we can crop it. We cannot crop it using the template raster yet because the crs' are different and cannot alter the template without affecting the grid. 
target_crs <- crs(canopy)
bc_canopy_boundary <- sf::st_transform(bc, target_crs)

# Crop canopy cover raster to bc bounding box
canopy_crop <- crop(canopy, bc_canopy_boundary) 

# Now that the layer is cropped to BC (rather than all of Canada), we can reproject it and mask it to our template
# Method = bilinear - use for continuous data
canopy_proj <- project(canopy_crop, template, method = "bilinear") 

# Mask to BC administrative boundary
canopy_mask <- mask(canopy_proj, template) 

# Save to cleaned constraints folder
writeRaster(canopy_mask, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/cleaned/canopy_cleaned.tif")
```

Reclassify to 1/0 binary such that canopy cover > 40% is 0 (unsuitable) and canopy cover < 40% is 1 (suitable). 

```{r}
#| message: false
#| eval: false

# Create reclassification scheme 
# 0 = not suitable - within constraints
# 1 = suitable - not within constraints
m <- c(0, 40, 1,
       40, 100, 0)

# Create matrix
canopy_matrix <- matrix(m, ncol = 3, byrow = TRUE)

# Apply classification
canopy_binary <- terra::classify(canopy_mask, canopy_matrix)

# Save to reclassified folder
writeRaster(canopy_binary, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/canopy_binary.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Constraint 3: Dense Forests"
cons_cols <- c("#696969", "#228B22")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/canopy_binary.tif"), col=cons_cols)
```

#### 1.3.4 Agricultural Land Reserve / Soil Capability 

Data source: [ALR GIS Data](https://www.alc.gov.bc.ca/alr-maps/) 

Data source: 

The ALR is a provincial land use zone that prioritizes and protects agricultural land. The *whole* ALR will not be included in the constraints - rather, only portions of the ALR with soil capability classes of 1 and 2 will be within the restriction zone. This will protect the most arable land from development, while also promoting the promising concept of agrovoltaics. 

To clean the data isolate polygons with a soil capability class less than 3 then intercept soil capability data with the ALR layer. Rasterize remaining data into a binary and mask to our template.

```{r}
#| message: false
#| eval: false

soils <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Mask/Raw Data/50KAgCap.gdb")
alr <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Mask/Raw Data/ALR_Polygons/ALR_Polygons.shp")

# View fields
# CC_Label = capability class
head(soils)

# Create function that selects polygons that are at least 50% class 1, 2, OR 3. Creates a new column called pct_123. 
pct_class_123 <- function(label) {
  if (is.na(label)) return(NA_real_)
  
  m <- str_match_all(label, "(?:(\\d+):)?(\\d)[A-Za-z]*")[[1]]
  
  pct   <- ifelse(is.na(m[,2]), 100, as.numeric(m[,2]) * 10)
  class <- as.numeric(m[,3])
  
  sum(pct[class %in% c(1, 2, 3)])
}

# Apply function to soils sf object
prime_soils <- soils %>%
  mutate(pct_123 = sapply(CC_Label, pct_class_123)) %>%
  filter(!is.na(pct_123) & pct_123 >= 50) 

# Intersect prime soils layer with the ALR polygons to create a new layer showing where the ALR and the prime soils polygons overlap. 
alr_soils_int <- sf::st_intersection(alr, prime_soils)

# Convert to spatvect
alr_soils_vect <- vect(alr_soils_int)

# Rasterize layer and spatially conform to our template
# Set field to 0 (unsuitable) and background to 1 (suitable) so all non-road cells are assigned a value of 1
alr_soils_rast <- rasterize(alr_soils_vect, template, field = 0, background = 1)

# Mask to the template 
alr_soils_binary <- mask(alr_soils_rast, template)

# Save to reclassified folder
writeRaster(alr_soils_binary, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/alr_soils_binary.tif", overwrite = TRUE)
```

```{r}
#| echo: false
#| fig-cap: "Constraint 4: Prime Agricultural Land"
cons_cols <- c("#696969", "#228B22")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/alr_soils_binary.tif"), col=cons_cols)
```

#### 1.3.5 Protected and Conserved Areas

Data source: [Canadian Protected and Conserved Areas Database](https://www.canada.ca/en/environment-climate-change/services/national-wildlife-areas/protected-conserved-areas-database.html) 

Clean and rasterize data as usual, with the additional step of adding a 1km ecological buffer around Protected and Conserved Areas (PCA) polygons before rasterizing. 

```{r}
#| message: false
#| eval: false

pca <- sf::st_read("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/Data/Mask/Raw Data/ProtectedConservedArea_2024.gdb", layer = "ProtectedConservedArea_2024")

# Create a BC boundary layer in the projection of the PCA raster so that we can crop it. We cannot crop it using the template raster yet because the crs' are different and cannot alter the template without affecting the grid. 
target_crs <- crs(pca)
bc_pca_boundary <- sf::st_transform(bc, target_crs)

# Clip PCA to BC bounding box
pca_clip <- sf::st_intersection(pca, bc_pca_boundary) 

# Buffer 1000m
pca_buff <- sf::st_buffer(pca_clip, 1000)

# Make spatvect 
pca_buff_vect <- vect(pca_buff)

# Now that the layer is cropped to BC (rather than all of Canada), we can reproject it and mask it to our template
pca_proj <- project(pca_buff_vect, template)

# Rasterize into a binary, ensuring it spatially conforms to template
pca_rast <- rasterize(pca_proj, template, field = 0, background = 1)

# Mask to the template 
pca_binary <- mask(pca_rast, template)

# Save to cleaned constraints folder
writeRaster(pca_binary, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/pca_binary.tif")
```

```{r}
#| echo: false
#| fig-cap: "Constraint 5: Protected and Conserved Areas"
cons_cols <- c("#696969", "#228B22")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/reclassified/pca_binary.tif"), col=cons_cols)
```

#### 1.3.6 Archaeologically and Culturally Sensitive Areas

WAITING ON DATA

### 1.4 Land Constraints Mask 

Now that all data representing restriction zones are in binary raster format, we will easily combine them together into one master constraint mask to use on our final suitability surface. 

Logic: If all cells across all layers equals 0, cell in final mask will equal 0. If even one cell across all layers equal 1, cell in final mask will equal 1. 

We can use simple multiplication to accomplish this.

```{r}
#| message: false
#| eval: false

constraints <- (pca_binary *
                alr_soils_binary *
                canopy_binary *
                roads_binary *
                rails_binary *
                lcr_binary)

# Save
writeRaster(constraints, "/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/mask/constraints.tif")
```

```{r}
#| echo: false
#| fig-cap: "Land Constraints Mask"
cons_cols <- c("#696969", "#228B22")
plot(rast("/Users/kenziethomson/Library/CloudStorage/OneDrive-UBC/RA/SolarModel_Guide/data/constraints/mask/constraints.tif"), col=cons_cols)
```

## Step 2: Analytic Hierarchy Process

## Step 3: Weighted Sum Overlay

# Stage 2: Technical Potential
